### 2023-06-01 Manuscript/Draft1.Rmd
manuscript update

The Northeast United States continental shelf spans from the Outer Banks of North Carolina to the Gulf of Maine. The region covers over 250,000 km$^2$ of ocean, extending over 200 km from shore in the largest areas in New England to just 30 km off shore in the southern regions. This ecologically diverse region contains approximately 18,000 vertebrate marine species. Commercial fisheries have been an important part of local economies for centuries. In 2019, New England fisheries produced $22 billion in sales, which sustained over 200,000 jobs [@Econ]. Maintaining a healthy ecosystem is therefore vital to sustained ecological health and economic prosperity of the region.

@@

The NEFSC has conducted a bottom trawl survey since 1963 to support assessment and management of the fish and invertebrate populations in the region [@azarovitz1981brief; @politis2014northeast]. The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf. The survey has created a rich time series data set with many uses including species-specific habitat identification, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments and ultimately quota limits. The survey takes place twice each year- once in the spring and again in the fall. Most spatial analyses and  projections of future distributions typically assume a constant survey catchability and availability over time. For this reason, NOAA's survey design includes sampling each strata in approximately the same 3-4 week time period in each season.

Stock assessment scientists choose from a number of approaches to obtain abundance estimates that are derived from the survey data. Methods range from traditional design-based estimates to model-based estimates that vary in complexity. Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest. These methods do not account for spatial variation in samples and are not able to account for environmental influences on survey values. Model-based abundance estimates use statistical models to measure the relationship between response variables (such as presence or abundance) and predictor variables (such as environmental factors). Model-based estimators, such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM), help account for complex relationships between variables and can help overcome problems with sampling design.

@@

Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades. The changes have resulted in a shift in spatial distributions of many species [@nye2009changing; @kleisner2017marine; @henderson2017effects]. Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments. More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey will change, altering the relationship between the index and the true population [@arreguin1996catchability; @langan2021climate]. A species shifting its range beyond the survey area is an additional compounding factor to consider. Existing research has focused on temperature as the driver of such changes [@klein2017effects] and evidence suggests that failing to account for the impact of climate-induced change can lead to management challenges [@kerr2022coherence]. In these scenarios, management strategy evaluations have shown that unintended overfishing can occur resulting from misconceptions of stock status, which can ultimately have detrimental ecologic and economic impacts [@mazur2023consequences]. We are therefore interested in analyzing the impact of climate change on the accuracy of abundance estimates derived from NOAA’s ongoing bottom-trawl survey along the East coast.

@@

We use the R package *MixFishSim* (MFS) to model our populations [@dolder2020highly]. MFS is a discrete spatiotemporal simulation tool where users can model multiple species under varying environmental conditions. The package uses a delay-difference population model with discrete processes for growth, death, and recruitment of the population. We formulate the following inputs for the MFS package to address our research question.

@@

We obtained a shapefile for the 15 strata that comprise Georges Bank, where strata were partitioned based primarily on depth and secondarily by latitude [@politis2014northeast]. The region was discritized into a raster with 88 rows and 144 columns to use as our modeling environment. Each cell in our simulation domain represents approximately 8.7 km$^2$. A fish stock is considered to be a subpopulation of a species that has similar intrinsic parameters. Each of the species being modeled has multiple distinct stocks along the Atlantic coast resulting from local environmental conditions. Biological differences between species results in each stock inhabiting a different number of strata on Georges Bank. Haddock inhabit all 15 strata in the domain, Cod populate 13 strata, and Yellowtail can be found in 9 strata. Figure \@ref(fig:strata-plot) shows the regions used in our models.

@@

The time step for our models is one week. MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass in each cell in our study area [@dolder2020highly]. Individual terms in the formulation account for growth of mature adults, mortality (natural and fishing), and the addition of new recruits. Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period. Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics. A full list of parameters used in our model can be seen below in Tables 1 and \@ref(tab:paramsSCENARIOS).

@@

The package was designed to generate hypothetical temperature gradients and theoretical habitat preferences using Gaussian Random Fields. The following sections describe how we formulated the habitat and temperature components to model real species in the western Atlantic Ocean.

@@

Species-specific habitat preferences were derived from niche model for each species using the *lrren* tool from the R package *envi* [@envi]. The *lrren* tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors. We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow. We combined data from both the fall and spring surveys to obscure the influence of temperature to allow the niche model to instead infer habitat preferences. Depth and mean sediment size were used as our covariate predictors. Estimated depth for the region was obtained from FVCOM [@chen2006unstructured]. The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method using point data collected by the United States Geologic Survey (USGS) [@mcmullen2005usgs]. Since the values in $Hab_{J,s}$ are required to be between 0 and 1, we rescaled the spatial estimates from *lrren* to fall between these bounds. See Figure \@ref(fig:hab-plot1) for a visual representation of this process being applied to Cod. Figure \@ref(fig:hab-plot) depicts habitat preferences $Hab_{J,s}$ for each species.

@@

Each species is assumed to have normally distributed temperature preferences ($N(\mu,\sigma)$). Values were chosen by combining information in the literature with temperatures recorded in the bottom trawl survey. We assume Yellowtail Flounder's preferences are $N(8.75,4.25)$, while Haddock and Cod have preferences $N(9,4)$. Weekly estimated temperature data for the region for 2012 was obtained from FVCOM [@chen2006unstructured]. We chose to repeat temperature estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data. The 2012 data was chosen because it displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values, allowing for a smooth repeating yearly temperature pattern for the constant temperature scenario. The 2012 temperature data was also transformed to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation. We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on dynamics while remaining within reasonable computational limits in terms of the length of the simulation. Figure \@ref(fig:temp-scenarios) depicts mean trends for the temperature scenarios used in our models. **dont forget to include animated gif in final submission**

@@

We carry out 20 year simulations for each of our three species under various population scenarios. Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year [[can cite the 2022 management track assessments, see https://apps-nefsc.fisheries.noaa.gov/saw/sari.php for when the document becomes available]] **cite**. For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock. To provide a comprehensive analysis of population indexing methods we consider all possible scenario combinations for Yellowtail Flounder. Each of these scenarios is simulated twice: first with with an oscillating temperature gradient that repeats and second with a temperature gradient that increases roughly 5 degrees Celsius over the duration of the 20 year simulation, for a total of 10 simulated spatial datasets. The specific population trends used in our analyses can be see in Figure \@ref(fig:pop-scenarios).

@@

After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year. We sample each strata in the same weeks in which the Spring and Fall surveys take place (weeks 13 & 14 in the Spring and 37 & 38 in the Fall). The number of the samples taken reflect true target values for each strata and sampling cells were randomly selected. We then use the biomass collected from our samples in contemporary abundance indexing methods to estimate population trends. Knowing the true population values in our simulations allows us to compute and compare the absolute error calculated from each estimation method.

@@

\[
  \makebox[\linewidth]{$SM_{s,y} = \sum_{i=1}^N W_i \frac{\sum_{k=1}^{S_i}y_{k,i}}{S_i}$},
\]

@@

\[
  \makebox[\linewidth]{$p_1(i)=\text{temporal variation} + \text{spatial variation} + \text{spatio-temporal variation}  + \text{habitat covariates} + \text{catchability covariates}$},
\]

\[
  \makebox[\linewidth]{$p_*(i)= \beta_*(t_i) + L_{\omega_*}\omega_*(s_i) + L_* \epsilon_*(c_i,t_i) + \upsilon(s_i,t_i) + \xi(i)$},
\]

where more specific functional forms of each component are further described in [@thorson2019guidance]. $p_2(i)$ takes on a similar form as $p_1(i)$. VAST models require numerous user inputs to determine how the linear predictors will be conditioned and solved. As a result, VAST models take on the order of hours to complete.

We follow the advice given in [@thorson2019guidance] to build VAST models that estimate biomass in our Georges Bank population models using stratified random samples from our model output. In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation. We carried out the same model-building process using covariate information to inform our models as well as without covariates in our models. The covariates we considered were the dynamic temperature values and static habitat values ($Hab_{J.s}$) from our population models. When using covariates we ultimately decided to consider a best-case scenario by including both temperature and habitat covariates for both linear predictors in order to provide the most information to the model. Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings. Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models, which can be seen in Table 3. VAST Settings B were chosen after reviewing the information provided in the "Seasonal Model" section of VAST's online Github tutorial [@ThorsonSeasonal]. After running models for each of our scenarios using these settings, we shared some of our results with the VAST package creator, James (Jim) Thorson. Jim reviewed our VAST model results and suggested we also consider the options shown in Settings A. Both settings use a Tweedie model and turn off the first linear predictor $p_1$ so that only biomass is being modeled. The difference between the two settings can be seen in the Rho Configuration, where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A. The difference amounts to intercepts for the biomass predictor being a fixed effect (RhoBeta2=0) versus intercepts being constant among years as a fixed effect (RhoBeta2=3). According to Jim, the Settings A values are representative of what stock assessment scientists typically use in their analyses. As a result we will focus our analysis on VAST results obtained using Settings A.

Our goal is to determine indexing approaches and settings that are robust to changing environmental conditions and resulting spatial biomass patterns. An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists. These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata. To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values. Annual survey observations were simulated as log-normal deviations from the underlying “true” survey catches with a CV of 0.3 in the spring. The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population. We therefore simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species. We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from. Figure \@ref(fig:strata-plot) shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \@ref(fig:PopPct). The yellow regions in Figure \@ref(fig:strata-plot) depict the strata that were removed from certain stratified random surveys for each species.

<!-- Note the Table XXX needs to be added comment above. -->
Each scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, covariates), resulting in a large number of scenario combinations to consider. The columns in Table \@ref(tab:scenarios) show the choices that define each scenario.

@@

Figure \@ref(fig:PopPct) depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey in the spring (weeks 13 and 14) and fall (weeks 37 and 38). The left column in Figure \@ref(fig:PopPct) depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern. In these scenarios we notice a small amount of shifting between successive years as the population aggregates on especially suitable habitat in the domain. More exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time, as seen right column of Figure \@ref(fig:PopPct). This demonstrates how we were able to model shifting distributions due to climate change by running simulations with a temperature gradient that increased on average over time.

Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the absolute error between abundance estimates and model output for each of our abundance estimates. While the model-based results provided a slightly lower mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02) highlighting the sensitivity of model-based estimates to the settings being used. When considering individual scenarios, $77/80\approx 96\%$ had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, with VAST models that include covariate information providing the lowest overall errors and standard deviations. Setting misspecification and/or a poor covariate response contribute to the large variance in VAST errors as illustrated by the fact that $\approx 63\%$ of individual scenarios contain a VAST estimate with a worst error than the corresponding stratified mean estimate. VAST models that included covariates had an average absolute error of 0.22, models that did not include covariate information had an average error of 0.46, and stratified mean estimates produces an average error of 0.39. When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal. Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain (e.g. in Table \@ref(tab:HadERROR) rows 8 and 9, VAST models without covariates show lower errors than when covariates are included). We analyze these results further in the sections that follow.

@@

A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values. This can be further examined by examining yearly estimate:model ratio values, where we divide the yearly abundance estimates by the true model value. In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can estimate yearly abundance values as low as 0 at times and exhibit large yearly changes. A representative example of this trend for each species is shown on the log scale in Figure \@ref(fig:RatioTrend).

Analysis of individual yearly model:estimate ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21. On the other hand, just 33\% of stratified mean estimate ratios were above 1 with an average of 0.874 and a standard deviation of 0.12. There were seasonal differences in estimate ratios for VAST with spring VAST ratios producing a mean value of 1.08 with a standard deviation of 0.12 while fall VAST ratios were larger with a mean of 1.50 and standard deviation of 0.38. Stratified mean ratios were more consistent between seasons with spring ratios resulting in a mean value of 0.91 with a standard deviation of 0.10, and fall values of providing a mean ratio of 0.84 with a standard deviation of 0.16. The breakdown for individual species followed a similar pattern.

@@

Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata) contain the absolute error between our abundance estimates for Yellowtail Flounder comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B). In reviewing these Tables we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors. The settings used in VAST mattered with Settings B outperforming Settings A in $67/96\approx 70\%$ of scenarios. When all strata are sampled, adding covariates greatly improved estimates in all scenarios. We still see an improvement in VAST estimates when certain strata are excluded from sampling, but the improvement was much less dramatic than with all strata included. There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata). However, Including covariate information to each VAST model produced estimates with significantly lower error than their stratified mean counterparts.

@@

In the constant temperature scenario for Cod, in both seasons the population decreases its presence in strata 19 and 20 over the duration of the simulation, while simultaneously increasing presence in stratum 16. However, we see a seasonal impact in stratum 16 during the increasing average temperature simulations where in the fall the population decrease presence in strata 16 and 21, and increase presence in 18, 22, and 24 (see Figure \@ref(fig:hab-plot)). Similar to the Yellowtail Flounder population, the favorable habitat in stratum 16 acts as an attractor in both temperature scenarios in the spring when the water temperature is cooler. When the temperature increases over time, the fall population compensates by shifting their preference to the adjacent strata that are deeper and/or further north than stratum 16.

>>>>

The Northeast United States continental shelf spans from the Outer Banks of North Carolina to the Gulf of Maine. The region covers over 250,000 km$^2$ of ocean, extending over 200 km from shore in the largest areas in New England to just 30 km off shore in the southern regions. This ecologically diverse region contains approximately 18,000 vertebrate marine species. Commercial fisheries have been an important part of local economies for centuries. In 2019, New England fisheries produced $22 billion in sales, which sustained over 200,000 jobs [@Econ]. Maintaining a healthy ecosystem is therefore vital to sustained ecosystem health and economic prosperity of the region.

@@

The Northeast Fisheries Science Center (NEFSC) has conducted a bottom trawl survey since 1963 to support assessment and management of the fish and invertebrate populations in the region [@azarovitz1981brief; @politis2014northeast]. The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf. The survey has created a rich time series data set with many uses including species-specific habitat identification, age and life history information, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments, and ultimately fishing quota limits. The survey takes place twice each year- once in the spring and again in the fall. Most spatial analyses and  projections of future distributions typically assume a constant survey catchability and availability over time. For this reason, NOAA's survey design includes sampling each strata in approximately the same 3-4 week time period in each season.

One component of a stock assessment is an estimation of yearly abundance that is often derived from the bottom trawl survey, when applicable. Stock assessment scientists choose from a number of approaches to obtain abundance estimates ranging from traditional design-based estimates to model-based estimates that vary in complexity. Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest. These methods do not account for spatial variation in samples and are not able to account for environmental influences on survey values. Model-based abundance estimates use statistical models to measure the relationship between response variables (such as presence or abundance) and predictor variables (such as environmental factors). Model-based estimators help account for complex relationships between variables and can help overcome problems with sampling design. Common model-based approaches include General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM).

@@

Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades. The changes have resulted in a shift in spatial distributions of many species [@nye2009changing; @kleisner2017marine; @henderson2017effects]. Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments. More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey could change, altering the relationship between the index and the true population [@arreguin1996catchability; @langan2021climate]. A species shifting its range beyond the survey area is an additional compounding factor to consider. Existing research has focused on temperature as the driver of such changes [@klein2017effects] and evidence suggests that failing to account for the impact of climate-induced change can lead to management challenges [@kerr2022coherence]. In these scenarios, management strategy evaluations have shown that misconceptions of stock status can lead to unintended overfishing, which can ultimately have detrimental ecologic and economic impacts [@mazur2023consequences]. We are therefore interested in analyzing the impact of climate change on the accuracy of abundance estimates derived from NOAA’s ongoing bottom-trawl survey along the East coast.

@@

We use the R package *MixFishSim* (MFS) to model our populations [@dolder2020highly]. MFS is a discrete spatiotemporal simulation tool that allows users to model multiple species under varying environmental conditions. The package uses a delay-difference population model with discrete processes for growth, death, and recruitment of the population. We formulate the following inputs for the MFS package to address our research question.

@@

We obtained a shapefile for the 15 strata that comprise Georges Bank, where strata are partitioned based primarily on depth and secondarily by latitude [@politis2014northeast]. The region was discritized into a raster with 88 rows and 144 columns to use as our modeling environment. Each cell in our simulation domain represents approximately 8.7 km$^2$. A fish stock is considered to be a subpopulation of a species that has similar intrinsic parameters. Each of the species being modeled has multiple biologically distinct stocks along the Atlantic coast resulting from local environmental conditions. As a result, each stock inhabits a different number of strata on Georges Bank. Haddock inhabit all 15 strata in the domain, Cod populate 13 strata, and Yellowtail can be found in 9 strata. Figure \@ref(fig:strata-plot) shows the strata used in our model for each species.

@@

The time step for our models is one week. MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass in each cell in our study area [@dolder2020highly]. Individual terms in the formulation account for growth of mature adults, mortality (natural and fishing), and the addition of new recruits. Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period. Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics. A full list of parameters used in our model can be seen in Tables 1 and \@ref(tab:paramsSCENARIOS).

@@

The package was designed to generate hypothetical temperature gradients and theoretical habitat preferences using Gaussian Random Fields. The following sections describe how we formulated the habitat and temperature components to model our 3 real species on Georges Bank.

@@

Species-specific habitat preferences were derived from niche model for each species using the *lrren* tool from the R package *envi* [@envi]. The *lrren* tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors. We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and by representing a successful catch by the biomass of the given tow. We combined data from both the fall and spring surveys to obscure the influence of temperature to allow the niche model to instead infer static habitat preferences. Depth and mean sediment size were used as our covariate predictors. Estimated depth for the region was obtained from FVCOM [@chen2006unstructured]. The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method using point data collected by the United States Geologic Survey (USGS) [@mcmullen2005usgs]. Since the values in $Hab_{J,s}$ are required to be between 0 and 1, we rescaled the spatial estimates from *lrren* to fall between these bounds. See Figure \@ref(fig:hab-plot1) for a visual representation of this process being applied to Cod. Figure \@ref(fig:hab-plot) depicts habitat preferences $Hab_{J,s}$ for each species.

@@

Each species is assumed to have normally distributed temperature preferences $N(\mu,\sigma)$, with mean $\mu$ and standard deviation $\sigma$. Values were chosen by combining information in the literature with temperatures recorded in the bottom trawl survey. We assume Yellowtail Flounder's preferences are $N(8.75,4.25)$, while Haddock and Cod have preferences $N(9,4)$. Weekly estimated temperature data for the region for 2012 was obtained from FVCOM [@chen2006unstructured]. We chose to repeat temperature estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data. The 2012 data was chosen because it displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values, allowing for a smooth repeating yearly temperature pattern for the constant temperature scenario. The 2012 temperature data was also transformed to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation. We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on dynamics while remaining within reasonable computational limits in terms of the length of the simulation. Figure \@ref(fig:temp-scenarios) depicts mean trends for the temperature scenarios used in our models. **dont forget to include animated gif in final submission**

@@

We carry out 20 year simulations for each of our three species under various population scenarios. Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year [[can cite the 2022 management track assessments, see https://apps-nefsc.fisheries.noaa.gov/saw/sari.php for when the document becomes available]] **cite**. For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock. To provide a comprehensive analysis of population indexing methods we consider all possible population trend combinations for Yellowtail Flounder. The specific population trends used in our analyses can be see in Figure \@ref(fig:pop-scenarios). Each of these scenarios is simulated twice: first with with an oscillating temperature gradient that repeats and second with a temperature gradient that increases roughly 5 degrees Celsius over the duration of the 20 year simulation, for a total of 10 simulated spatial datasets.

@@

After each simulation is complete, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year. We sample each strata in the same weeks in which the Spring and Fall surveys take place (weeks 13 & 14 in the Spring and 37 & 38 in the Fall). The number of the samples taken reflect true target values for each strata and sampling cells were randomly selected. 
An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists. These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata. To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values. Annual survey observations were simulated as log-normal deviations from the underlying “true” survey catches with a CV of 0.3 in the spring. The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population. We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species. The strata to exclude for each species were chosen by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from. Figure \@ref(fig:strata-plot) shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \@ref(fig:PopPct). The yellow regions in Figure \@ref(fig:strata-plot) depict the strata that were removed from certain stratified random surveys for each species. We then use the biomass collected from our samples in contemporary abundance indexing methods to estimate population trends. Knowing the true population values in our simulations allows us to compute and compare the absolute error calculated from each estimation method.

@@

\[
  \makebox[\linewidth]{$SM_{s,y} = \sum_{i=1}^N W_i \frac{\sum_{k=1}^{S_i}y_{k,i}}{S_i},$}
\]

@@

<!-- \[ -->
<!--   \makebox[\linewidth]{$p_1(i)=\text{temporal variation} + \text{spatial variation} + \text{spatio-temporal variation}  + \text{habitat covariates} + \text{catchability covariates}$}, -->
<!-- \] -->

\[
  \makebox[\linewidth]{$p_*(i)= \beta_*(t_i) + L_{\omega_*}\omega_*(s_i) + L_* \epsilon_*(c_i,t_i) + \upsilon(s_i,t_i) + \xi(i),$}
\]

for $*=1,2$, where more specific functional forms of each component are further described in [@thorson2019guidance]. VAST models require numerous user inputs to determine how the linear predictors will be conditioned and solved. As a result, VAST models take on the order of hours to complete.

We follow the advice given in [@thorson2019guidance] and [@VASTwiki] to build VAST models that estimate biomass in our Georges Bank population models using stratified random samples from our model output. In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation. We carried out the same model-building process using covariate information to inform our models as well as without covariates in our models. The covariates we considered were the driving factors of movement in our population models- the dynamic temperature values and static habitat values ($Hab_{J.s}$). When using covariates we ultimately decided to provide the most information to the model by including both temperature and habitat covariates to consider a best-case scenario. Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings. Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models as shown in Table 3. VAST Settings B were chosen after reviewing the information provided in the "Seasonal Model" section of VAST's online Github tutorial [@ThorsonSeasonal]. After running models for each of our scenarios using these settings, we shared some of our results with the VAST package creator, James (Jim) Thorson. Jim reviewed our VAST model results and suggested we also consider the options shown in Settings A. Both settings use a Tweedie model and turn off the first linear predictor $p_1$ so that only biomass is being modeled. The difference between the two settings can be seen in the Rho Configuration, where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A. The difference amounts to intercepts for the biomass predictor being a fixed effect (RhoBeta2=0) versus intercepts being constant among years as a fixed effect (RhoBeta2=3). According to Jim, the Settings A values are representative of what stock assessment scientists typically use in their analyses. As a result we will focus our analysis on VAST results obtained using Settings A.

Each scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, covariates), resulting in a large number of scenario combinations to consider. The columns in Table \@ref(tab:scenarios) show the choices that define each scenario.

@@

Figure \@ref(fig:PopPct) depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey in the spring (weeks 13 and 14) and fall (weeks 37 and 38). The left column in Figure \@ref(fig:PopPct) depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern. We notice a small amount of shifting between successive years in these constant temperature scenarios as the population aggregates on especially suitable habitat in the domain. As seen right column of Figure \@ref(fig:PopPct), more exaggerated shifting takes place in a larger number of strata when the temperature is increasing over time. By running simulations in *MixFishSim* with a temperature gradient that increased on average over time we were able to create spatial datasets with shifting biomass distributions.

Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the absolute error between biomass estimates and model output for each of our abundance estimates. While the model-based results provided a slightly lower overall mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02). When considering individual scenarios, while $77/80\approx 96\%$ had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, $63\%$ of individual scenarios contained a VAST estimate with a worst error than the corresponding stratified mean estimate. VAST models that included covariate information provided the lowest overall errors and standard deviations. Models with covariates had an average absolute error of 0.22, compared to models that did not include covariate information having an average error of 0.46, and stratified mean estimates produces an average error of 0.39. These findings demonstrate the ability of spatio-temporal models to account for spatial variation in the species being sampled under the right conditions, but also highlight volatility/sensitivity of model estimates to the settings being used and/or the covariate response detected in the data.

When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal. Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain (e.g. in Table \@ref(tab:HadERROR) rows 8 and 9, VAST models without covariates show lower errors than when covariates are included). We analyze these results further in the sections that follow.

@@

A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values. This can be further examined through yearly estimate:model ratio values, where we divide the yearly abundance estimates by the true model value. In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can result in yearly abundance values as low as 0 at times and exhibit large yearly changes. A representative example of this trend for each species is shown on the log scale in Figure \@ref(fig:RatioTrend).

Analysis of individual yearly model:estimate ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21. On the other hand, just 33\% of stratified mean estimate ratios were above 1 with an average of 0.874 and a standard deviation of 0.12. There were seasonal differences in estimate ratios for VAST with spring VAST ratios producing a mean value of 1.08 with a standard deviation of 0.12 while fall VAST ratios were larger with a mean of 1.50 and standard deviation of 0.38. Stratified mean ratios were more consistent between seasons with spring ratios resulting in a mean value of 0.91 with a standard deviation of 0.10, and fall values of providing a mean ratio of 0.84 with a standard deviation of 0.16. The breakdown for individual species followed similar patterns.

@@

Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata) contain the absolute error between our abundance estimates for Yellowtail Flounder comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B). In reviewing these Tables we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors. The settings used in VAST mattered with Settings B outperforming Settings A in $67/96\approx 70\%$ of scenarios. When all strata are sampled, adding covariates improved estimates in all scenarios. Including covariates still improve VAST estimates when certain strata are excluded from sampling, but the change in error was smaller than with all strata included. There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata), but including covariate information to each VAST model produced estimates with significantly lower error than the corresponding stratified mean estimate

@@

In the constant temperature scenario for Cod, the population decreases its presence in strata 19 and 20 in both seasons over the duration of the simulation, while simultaneously increasing presence in stratum 16. However, we see a seasonal impact in stratum 16 during the increasing average temperature simulations where in the fall the population decrease presence in strata 16 and 21, and increase presence in 18, 22, and 24 (see Figures \@ref(fig:strata-plot) and \@ref(fig:PopPct)). Similar to the Yellowtail Flounder population, the favorable habitat in stratum 16 acts as an attractor in both temperature scenarios in the spring when the water temperature is cooler. When the temperature increases over time, the fall population compensates by shifting their preference to the adjacent strata that are deeper and/or further north than stratum 16.
